Supplementary Information

### 2021-08-10 doc/manuscript/ForestGEO_treerings.Rmd
fixes to use of "linear"
#138

Tree rings provide an invaluable long-term record for understanding how climate and other drivers shape tree growth and forest productivity.
However, conventional tree-ring analysis methods were not designed to simultaneously test effects of climate, tree size, and other drivers on individual growth.
This has limited the potential to test ecologically relevant hypotheses on tree growth sensitivity to multiple climatic drivers and their interactions with tree size.
Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually maximum temperature, over $\le$ 3-month seasonal windows), with non-linear responses prevalent (63% of relationships).
Climate sensitivity commonly varied with DBH (44% of cases tested), with larger trees generally showing greater sensitivity. 
Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently reached a maximum at intermediate DBH.
Accounting for climate and DBH, growth rate declined over time for 92% of species in secondary or disturbed stands, whereas growth trends were mixed in older forests.
These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing growth than has previously been possible.

@@

To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of individual trees within a stand based on the primary biotic and abiotic drivers.
Specifically, what is needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter at breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree-ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
Below, we outline major hypotheses regarding the influence of these factors on tree growth that can best be addressed using a multifactorial, non-linear approach to tree-ring analysis (Table `r hypotheses`). 
We then develop such a framework and apply it to test these hypotheses.

@@

For RW, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear DBH term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative). 
There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
On one end of the spectrum, some species exhibited maximum RW at low DBH, followed by fairly rapid declines in RW with increasing DBH. 
This pattern was common among light-demanding species (6 of 13 site-species combinations; Tables `r hypotheses`, `r species_table`; e.g., *Melia azedarach* at HKK, *Populous tremuloides* at CB) and in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig. `r size`).
At the other end of the spectrum, some species had low RW at small DBH, increased to peak RW at intermediate DBH, and subsequently declined.
This pattern was common among shade-tolerant species (8 of 11 site-species combinations; Table `r hypotheses`; e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp. at SCBI and ZOF, *Picea* spp. at ZOF and CB; Table `r species_table`).

@@

Our analysis differed fundamentally from most conventional approaches in testing for non-linear responses of growth to climate, finding that the majority of climate responses were nonlinear (Table `r hypotheses`, Fig. `r clim_sens`). 
This result, which is consistent with physiological expectations [e.g., @kumarathunge_acclimation_2019; @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is approximately linear. 
The nonlinear form of most climate growth responses implies that as the climate changes such that high temperatures and strong precipitation anomalies become more common [@ipcc_climate_2014], non-stationary climate responses, already common [@wilmking_global_2020], could become more prevalent [@germain_climate_2020].

@@

Within the three older forests (ZOF, LT, CB), population-mean growth trends were mixed (Table `r hypotheses`, Fig. `r year_fig`), probably reflecting some combination of successional changes, changing mortality rates, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
In particular, light-demanding species that establish in gaps (e.g., *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s, consistent with other studies from central Europe showing dramatic growth reductions due to acid deposition during this period [@elling_dendroecological_2009; @samonil_longterm_2008]. 
Non-linear trends such as this would be more accurately described by a non-linear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
Notably, there were only three species -- *Fagus sylvatica* at ZOF and *Picea pungens* and *Pinus flexilis* at CB -- whose population-mean growth rate increased significantly over the analysis time frame (Fig. `r year_fig`).

>>>>

Tree rings provide an invaluable long-term record for understanding how climate and other drivers shape tree growth and forest productivity.
However, conventional tree-ring analysis methods were not designed to simultaneously test effects of climate, tree size, and other drivers on individual growth.
This has limited the potential to test ecologically relevant hypotheses on tree growth sensitivity to multiple climatic drivers and their interactions with tree size.
Here, we develop and apply a new method to simultaneously model nonlinear effects of primary climate drivers, reconstructed tree diameter (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually maximum temperature, over $\le$ 3-month seasonal windows), with concave-down responses in 63% of relationships.
Climate sensitivity commonly varied with DBH (44% of cases tested), with larger trees generally showing greater sensitivity. 
Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently reached a maximum at intermediate DBH.
Accounting for climate and DBH, growth rate declined over time for 92% of species in secondary or disturbed stands, whereas growth trends were mixed in older forests.
These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing growth than has previously been possible.

@@

To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of individual trees within a stand based on the primary biotic and abiotic drivers.
Specifically, what is needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter at breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
While multifactorial and sometimes nonlinear individual-based analysis frameworks have been applied in tree-ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
Below, we outline major hypotheses regarding the influence of these factors on tree growth that can best be addressed using a multifactorial, nonlinear approach to tree-ring analysis (Table `r hypotheses`). 
We then develop such a framework and apply it to test these hypotheses.

@@

For RW, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative). 
There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
On one end of the spectrum, some species exhibited maximum RW at low DBH, followed by fairly rapid declines in RW with increasing DBH. 
This pattern was common among light-demanding species (6 of 13 site-species combinations; Tables `r hypotheses`, `r species_table`; e.g., *Melia azedarach* at HKK, *Populous tremuloides* at CB) and in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig. `r size`).
At the other end of the spectrum, some species had low RW at small DBH, increased to peak RW at intermediate DBH, and subsequently declined.
This pattern was common among shade-tolerant species (8 of 11 site-species combinations; Table `r hypotheses`; e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp. at SCBI and ZOF, *Picea* spp. at ZOF and CB; Table `r species_table`).

@@

Our analysis differed fundamentally from most conventional approaches in testing for nonlinear responses of growth to climate, finding that the majority of climate responses were nonlinear (Table `r hypotheses`, Fig. `r clim_sens`). 
This result, which is consistent with physiological expectations [e.g., @kumarathunge_acclimation_2019; @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is approximately linear. 
The nonlinear form of most climate growth responses implies that as the climate changes such that high temperatures and strong precipitation anomalies become more common [@ipcc_climate_2014], non-stationary climate responses, already common [@wilmking_global_2020], could become more prevalent [@germain_climate_2020].

@@

Within the three older forests (ZOF, LT, CB), population-mean growth trends were mixed (Table `r hypotheses`, Fig. `r year_fig`), probably reflecting some combination of successional changes, changing mortality rates, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
In particular, light-demanding species that establish in gaps (e.g., *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s, consistent with other studies from central Europe showing dramatic growth reductions due to acid deposition during this period [@elling_dendroecological_2009; @samonil_longterm_2008]. 
Nonlinear trends such as this would be more accurately described by a nonlinear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
Notably, there were only three species -- *Fagus sylvatica* at ZOF and *Picea pungens* and *Pinus flexilis* at CB -- whose population-mean growth rate increased significantly over the analysis time frame (Fig. `r year_fig`).
